---
title: "01_data_manipulation"
author: "VC"
date: "6/28/2020"
output: html_document
---



Why use tidyverse?
1. Base R can be painful with different data formats
2. Its more intuitive and useful for new learners

For example, often base R forces character to be converted to factors


##dplyr package 
 # dplyr package makes tabular data manipulation easier
 # Many functions are compiled in C++ so its faster
 # You can work with data stored in external data bases like SQL so removing limitation of memory as an issue
 
##tidyr package 
 # tidyr allows easier conversion between different data formats e.g., long and wide
 #It makes such conversion easy

Both these packages are available in umbrella package Tidyverse


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data


```{r }
data<- download.file(url="https://ndownloader.figshare.com/files/2292169",
              destfile = "data/combined.csv")
surveys<- read_csv("data/combined.csv")
head(surveys)
```

## Packages


```{r}
#install.packages("tidyverse")
library(tidyverse)
```

## Packages

```{r}
str(surveys)
#View(surveys)

```
# Tibble
 Tibble is like a data frame, characters are not coerced into columns
 
# Functions we will use

1. Select: subset columns
2. Filer: subset rows based on condition
3. mutate: create new columns by using info from other column
4. group_by() and summarize(): create summary statistics on grouped data
5. arrange(): sort results
6. count(): count discrete values

# Select specific columns
```{r}
select(surveys, plot_id,species_id,weight)
```
# Select all columns but other than specific columns
```{r}
select(surveys, -record_id, -species_id)
```
# Select rows based on specific condition
```{r}
filter(surveys, year==1995)
```

# Doing multiple steps 
```{r}
#select weight > 5g
surveys2<-filter(surveys,weight<5) 
# now slect only few columns
surveys2_sml<- select(surveys2,species_id,sex,weight )
```

# We can do the same by nesting functions
```{r}
#select weight > 5g
surveys2_sml<-select(filter(surveys, weight<5), species_id,sex,weight  )

```

# Doing multiple steps at a time
#mac cmd+shift+m
#windows ctrl+shift+m
```{r}
#select weight > 5g
surveys2<-surveys %>% filter(weight<5) %>% 
  select(species_id,sex,weight )
```

# Challenge
Using pipes, subset the surveys data to include animals collected before 1995 and retain only the columns year, sex, and weight.


